Tools for identifying unexpectedly low microfossil count sums

### 2019-05-18 Rmd/count_check_MS.Rmd
Cathy's edits

Several numerical tools have been developed to identify questionable data.
Deviations from the distribution of digits expected from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017].
Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
Brown et al. [-@Brown2017] developed the granularity-related inconsistency means (GRIM) test of whether means of integer data are consistent with the reported sample size.
However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often fifty for chironomids and several hundred for pollen and diatoms.
This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001].
However, mistakes happen and metadata such as count sums can be forgotten once percentages are calculated.
In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.
The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules; a subsequent corrigendum [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.

@@

In general, it is not possible to determine the sum from which percent were calculated, but the properties of community and assemblage counts make it possible to estimate it.
Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
This method will fail for assemblages without singletons.
We also expect the percent _p_ to be calculated from integer counts, therefore, there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.

@@

Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness.
Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded. 
Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
Datasets with possible mis-reporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
This paper does not attempt to be an exhaustive survey of all the data available.

@@

To test the sensitivity of the prevalence of counts without singletons to the count sum, I re-sample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
The prevalence of counts with singletons is low for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
With even larger counts the prevalence of counts with singletons decreases due to saturation as singletons become doubletons and few new taxa are found.

```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton orders. Results are the mean of ten trials, error bars are two standard errors.", fig.width = 70/25.4, fig.height = 70/25.4}
readd(bird_recount_singleton_plot)
```

The `r nrow(owens_summ1)` diatoms counts from Owen's Lake span a wide range of count sums (`r min(owens_summ1$count_sum)` -- `r max(owens_summ1$count_sum)`) as, if concentrations were low, the requirement that at least 300 valves were counted, was replaced by a minimum area of slide to count [@Platt1997].
Discounting the `r sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(. <= 10) %>% invisible()`), `r owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of `r owens_summ$median_sing %>% as.english()`.
Assemblages with count sums above 50 have a higher chance of having singletons (`r owens_summ$no_single_gt_50`% vs `r owens_summ$no_single50_`%).
`r owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of `r owens_summ$gcd_max %>% as.english()`.
Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.

@@

The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` thousand assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: `r pollen_summ$dataset_gcd1`% of the datasets have no such assemblages.
Some data sets with a high proportion of assemblages with a greatest common divisor above are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
For example, one dataset includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
The greatest common divisor is three for all assemblages, that is all `r pollenA$n_count` counts in the dataset are divisible by three.
Excluding such datasets would further increase the proportion of assemblages with singletons. 
Some other Neotoma datasets are discussed is a subsequent section.

@@

The Owen's Lake dataset [@Platt1997] also includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified. Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages. The few assemblages that fail the direct search method are species poor and have low count sums.

@@

```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums by the minumum percent method (solid symbols with error bars) and the direct search method (red crosses) from Last Chance Lake [@Axford2017]. Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 90/25.4, fig.height = 90/25.4}
readd(last_chance_plot)
```

@@

Some of the pollen data in the Neotoma database have an inexplicably high proportions of assemblages with a greatest common divisor greater than one. 
Here I focus on four datasets produced by the same research group. 
All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table))).

```{r neotoma-table}
pollen1234$table %>% knitr::kable(caption = "Caption")
```

Assemblages with a greatest common divisor of one typically have a higher taxonomic richness than assemblages with a greatest common divisor of two (Fig. \@ref(fig:neotoma-plot)).
The latter have a richness that is typically of the former with count sums half as large.

```{r neotoma-plot, fig.height = 100/24.5, fig.width=100/24.5, fig.cap="Number of taxa against count sum for assemblages with a greatest common divisor of one (red circles) or two (blue triangles). All four datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two."}
pollen1234$figure %>% print()
```

@@

Dataset diatom2 has `r diatom2_summ$nsamples` assemblages with a median richness of `r diatom2_summ$median_taxa` taxa.
The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
The direct search method estimates that `r diatom2_summ$n400 %>% as.english()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% as.english()` have less than 100.
The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
This means that for the assemblage with the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.

Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% as.english` assemblages had count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.

Palaeoceanographic dataset marine1 has `r marine1_summ$nsamples` assemblages with a median taxonomic richness of `r marine1_summ$median_taxa` taxa.
The associated paper reports that assemblages with count sums below 100 were omitted, and that there were 20 assemblages with count sums below 200.
The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
Some of the divergence appears to be because some, mainly low diversity, assemblages lack singletons.
The most extreme divergences appear to be data entry errors with incorrect rounding, perhaps during taxonomic revisions of the percent data rather than the raw counts.

@@

Analysis of the breeding bird, Owen's Lake diatom, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts have singletons, and an even larger proportion have a greatest common divisor of one. 
Therefore, these characteristics are potentially useful for identifying data where the count sum is misreported.

My search through archived microfossil data found several datasets where count sums are, or appear to be, smaller than that reported.
Some of the count sums appear to be an order of magnitude below what was reported.
The percent datasets examined are a convenience sample of available data that met the inclusion criteria.
As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
It is possible in some cases that the assemblages with low counts were omitted or merged prior to analysis, but the number of assemblages reported was not updated.
Care was taken to identify any such cases by, for example, examining stratigraphic diagrams to see if the low count assemblages were included.
In some cases, the assemblages with low apparent counts can be seen in the published stratigraphic diagrams.

It is also possible that the apparently low count sums are a false positive. 
However, as taxonomically diverse assemblages with a greatest common divisor greater than one are rare and all the datasets with apparently low count sums contained multiple taxonomically diverse assemblages with apparently low count sums, this is unlikely.

These tests require the assemblage data which are often not archived. 
However, if the count sums are very low it is possible to recognise this from a stratigraphic diagram. 
For example, a chironomid stratigraphy had an assemblage where all five taxa had a relative abundance of about 20%, suggesting a count sum of five, far below the reported fifty chironomids. Another assemblage had six taxa with a relative abundance of just over 10% and a seventh with twice as much, suggesting a count sum of eight chironomids.
On request, the authors sent the data. 
None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.

@@

The direct search method will fail if the dataset includes taxa calculated with different count sums, for example pollen sums of trees, shrubs and upland herbs and a pollen and spore sum that also includes pteridiophytes. 
It should be possible to identify such cases from the meta-data and knowledge of usual practice with different proxies. 
It will also fail, as show by dataset marine1, if percentages are incorrectly calculated or rounded.

Some taxonomic groups have microfossils that often come in groups of attached individuals.
For example, for each diatom cell has two valves which are the counting unit.
These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
This might make singletons slightly less likely. 
The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.

>>>>

Several numerical tools have been developed to identify questionable data.
Deviations from the distribution of digits expected from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017].
Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
Brown et al. [-@Brown2017] developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size.
However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often fifty for chironomids and several hundred for pollen and diatoms.
This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
However, mistakes happen and metadata such as count sums can be forgotten once percentages are calculated.
In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.
The risk that count sums will be misreported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules; a subsequent corrigendum [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.

@@

In general, it is not possible to determine the sum from which percent were calculated, but the properties of community and assemblage counts make it possible to estimate it.
Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
This method will fail for assemblages without singletons.
We also expect the percent _p_ to be calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.

@@

Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded. 
Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
This paper does not attempt to be an exhaustive survey of all the data available.

@@

To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
The prevalence of counts with singletons is low for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
With even larger counts the prevalence of counts with singletons decreases due to saturation as singletons become doubletons and few new taxa are found.

```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton at the order level. Results are the mean of ten trials, error bars are two standard errors.", fig.width = 80/25.4, fig.height = 70/25.4}
readd(bird_recount_singleton_plot)
```

The `r nrow(owens_summ1)` diatom counts from Owen's Lake span a wide range of count sums (`r min(owens_summ1$count_sum)` -- `r max(owens_summ1$count_sum)`) because, if concentrations were low, the requirement that at least 300 valves were counted was replaced by a minimum area of slide to count [@Platt1997].
Discounting the `r sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(. <= 10) %>% invisible()`), `r owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of `r owens_summ$median_sing %>% as.english()`.
Assemblages with count sums above 50 have a higher chance of having singletons (`r owens_summ$no_single_gt_50`% vs `r owens_summ$no_single50_`%).
`r owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of `r owens_summ$gcd_max %>% as.english()`.
Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.

@@

The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
For example, one dataset includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
The greatest common divisor is three for all assemblages, that is all `r pollenA$n_count` counts in the dataset are divisible by three.
Excluding such datasets would further increase the proportion of assemblages with singletons. 
Some other Neotoma datasets are discussed is a subsequent section.

@@

The Owen's Lake dataset [@Platt1997] includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified. Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages. 
The few assemblages that fail the direct search method are species poor and have low count sums.

@@

```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums by the minumum percent method (solid symbols with error bars) and the direct search method (red crosses) from Last Chance Lake [@Axford2017]. Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 80/25.4, fig.height = 80/25.4}
readd(last_chance_plot)
```

@@

Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
Here I focus on four datasets produced by the same research group. 
All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table)).

```{r neotoma-table}
pollen1234$table %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
```

Assemblages with a greatest common divisor of one typically have a higher taxonomic richness than assemblages with a greatest common divisor of two (Fig. \@ref(fig:neotoma-plot)).
In the datasets with a wide range of count sums, the richness of the assemblages with a greatest common divisor of two is comparable with the richness of the assemblages with a greatest common divisor of one that have a count sum that is half as large.

```{r neotoma-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles). Lines are loess with a span of 1. All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
pollen1234$figure %>% print()
```

@@

Dataset diatom2 has `r diatom2_summ$nsamples` assemblages with a median richness of `r diatom2_summ$median_taxa` taxa.
The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
The direct search method estimates that `r diatom2_summ$n400 %>% as.english()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% as.english()` have less than 100.
The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.

Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% as.english` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.

Palaeoceanographic dataset marine1 has `r marine1_summ$nsamples` assemblages with a median taxonomic richness of `r marine1_summ$median_taxa` taxa.
The associated paper reports that assemblages with count sums below 100 were omitted.
The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
Some of the divergence appears to be because some, mainly low diversity, assemblages lack singletons.
The most extreme divergences appear to be data entry errors with incorrect rounding, perhaps during taxonomic revisions of the percent data rather than the raw counts.

@@

Analysis of the breeding bird, Owen's Lake diatom, testate amoeba and pollen datasets has shown that, as expected, the vast majority of community and assemblage counts have singletons, and an even larger proportion have a greatest common divisor of one. 
Therefore, these characteristics are potentially useful for identifying data where the count sum is misreported.

My search through archived microfossil data found several datasets where count sums are, or appear to be, smaller than that reported.
Some of the count sums appear to be an order of magnitude below what was reported.
The non-pollen datasets examined here are a convenience sample of available data that met the inclusion criteria.
As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
It is possible in some cases that the assemblages with low counts were omitted or merged prior to analysis, but the number of assemblages reported was not updated.
Care was taken to identify any such cases by, for example, examining stratigraphic diagrams to see if the low count assemblages were included.
In some cases, the assemblages with low apparent counts can be seen in the published stratigraphic diagrams.

It is also possible that the apparently low count sums are a false positive. 
However, as taxonomically diverse assemblages with a greatest common divisor greater than one are rare and all the datasets with apparently low count sums contained several such assemblages, this is unlikely.

These tests require the assemblage data which are often not archived. 
However, if the count sums are very low it is possible to recognise this from a stratigraphic diagram. 
For example, a chironomid stratigraphy had an assemblage where all five taxa had a relative abundance of about 20%, suggesting a count sum of five, far below the reported fifty chironomids. Another assemblage had six taxa with a relative abundance of just over 10% and a seventh with twice as much, suggesting a count sum of eight chironomids.
On request, the authors sent the data. 
None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more that fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.

@@

The direct search method will fail if the dataset includes taxa calculated with different count sums, for example pollen sums of trees, shrubs and upland herbs and a pollen and spore sum that also includes pteridiophytes. 
It should be possible to identify such cases from the meta-data and knowledge of usual practice with different proxies. 
It will also fail, as shown by dataset marine1, if percentages are incorrectly calculated or rounded.

Some taxonomic groups have microfossils that often come in groups of attached individuals.
For example, each diatom cell has two valves which are the counting unit.
These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
This might make singletons slightly less likely. 
The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.
